\chapter{Object Ownership}

Qt has a parent-owned object model. Crystal has a garbage collector. \api{crystal-qt6} keeps those two worlds separate by making native Qt cleanup deterministic instead of depending on Crystal finalizers.

This chapter explains the two wrapper families in the \api{crystal-qt6} library, how ownership moves through parent relationships, when \api{release} matters, and which pitfalls to avoid when building larger applications.

\section{\api{QObject} Wrappers}

Most visible desktop objects in Qt inherit from \api{QObject}: widgets, timers, actions, menus, models, dialogs, and many support objects. In \api{crystal-qt6}, those classes inherit from \api{Qt6::QObject}.

A \api{QObject} wrapper stores a native handle, knows whether the Crystal wrapper owns that handle, and listens for Qt's native \api{destroyed} signal. When Qt destroys the native object, the wrapper marks itself as destroyed and stops participating in shutdown cleanup.

\begin{lstlisting}[style=crystal]
main = Qt6::MainWindow.new
button = Qt6::PushButton.new("Apply", main)

button.destroyed.connect do
  puts "button native object was destroyed"
end
\end{lstlisting}

The \api{destroyed?} predicate is mostly useful for defensive code and tests. Ordinary application code should be structured so it does not keep using widgets after their owning window has been torn down.

\section{Parent-Owned Objects}

Qt's basic ownership rule is simple: a parent destroys its children. The Crystal wrappers follow that rule. Constructors usually accept an optional parent. If you pass a parent, Qt owns the native object. If you omit the parent, the wrapper owns it until release or process shutdown.

\begin{lstlisting}[style=crystal]
main = Qt6::MainWindow.new

# Owned by the Crystal wrapper until Qt6.shutdown.
floating_timer = Qt6::QTimer.new

# Owned by main because main is passed as the parent.
window_timer = Qt6::QTimer.new(main)

# Owned by main after it is added to the window.
toolbar = Qt6::ToolBar.new("Main", main)
main.add_tool_bar(toolbar)
\end{lstlisting}

Top-level windows are normally wrapper-owned. Child widgets, actions, timers, menus, docks, layouts, and helper objects should usually be parented to a durable object such as the main window.

\section{Ownership At A Glance}

Most wrapper types fit into a small number of lifetime patterns:

\begin{longtable}{p{0.24\textwidth}p{0.66\textwidth}}
\toprule
Wrapper & Ownership pattern \\
\midrule
\api{Widget} and subclasses & Usually parented, layout-adopted, or wrapper-owned as a top-level window. Parent child widgets or add them to layouts. Keep top-level windows alive until the UI is finished. \\
\api{Action} & Usually parented to a \api{QObject}, menu, toolbar, or action group. Parent actions to the main window or let menus/toolbars adopt them. Do not release actions still installed in UI chrome. \\
\api{QTimer} & Usually parented to a \api{QObject} or wrapper-owned. Parent repeating timers to a durable object. Stop or release short-lived unparented timers when work is finished. \\
\api{QImage} and \api{QPixmap} & Wrapper-owned \api{NativeResource} objects. Use normally as values. Explicit release is only useful in tight loops or when freeing large rasters early matters. \\
\api{QPainter} & Wrapper-owned active paint-device helper. End the painter's lifetime before reading or saving the target device. Prefer block forms when available. \\
\api{Clipboard} & Borrowed process-wide object returned by \api{Qt6.clipboard}. Never release the returned wrapper. \\
\api{StatusBar} & Borrowed from the main window through \api{main.status_bar}. Never release the borrowed status bar wrapper. \\
\bottomrule
\end{longtable}

When in doubt, ask one question: who needs this native object to stay alive? If the answer is "the main window," give it a parent. If the answer is "only this short block of code," an explicit \api{ensure} release or a block helper is appropriate. If the object came from a getter such as \api{status_bar}, treat it as borrowed.

\section{\api{NativeResource} Wrappers}

Not every Qt type is a \api{QObject}. Value-like or device-like objects such as images, pixmaps, painters, paths, brushes, pens, fonts, transforms, dates, URLs, directories, and settings are represented by \api{Qt6::NativeResource}.

\api{NativeResource} is simpler than \api{QObject}. It tracks whether the wrapper owns a native handle, exposes \api{release}, and calls a type-specific native destructor. It does not have Qt parent ownership or a native \api{destroyed} signal.

\begin{lstlisting}[style=crystal]
image = Qt6::QImage.new(320, 180, Qt6::ImageFormat::ARGB32)
image.fill(Qt6::Color.new(245, 248, 252))

path = Qt6::QPainterPath.new
path.move_to(20, 20)
path.line_to(180, 80)
\end{lstlisting}

Most small applications can rely on the automatic shutdown path for these resources. Explicit release becomes more important when you create many short-lived native resources in a loop, when a file/device needs to be finalized before reading it back, or when an example intentionally demonstrates deterministic cleanup.

\section{Painter And Device Lifetime}

\api{QPainter} is a \api{NativeResource}, but it deserves special attention because it actively writes to another object. While a painter is active, Qt considers the target paint device in use. Release the painter before saving, reading, converting, or reusing that target.

\begin{lstlisting}[style=crystal]
image = Qt6::QImage.new(320, 180, Qt6::ImageFormat::ARGB32)
image.fill(Qt6::Color.new(255, 255, 255))

Qt6::QPainter.paint(image) do |painter|
  painter.pen = Qt6::Color.new(40, 80, 120)
  painter.draw_text(Qt6::PointF.new(24.0, 48.0), "Rendered")
end

# The painter has been released, so the image can now be saved.
image.save("preview.png")
\end{lstlisting}

If you create a painter manually, use \api{ensure}:

\begin{lstlisting}[style=crystal]
painter = Qt6::QPainter.new(image)
begin
  painter.draw_rect(Qt6::RectF.new(10.0, 10.0, 120.0, 80.0))
ensure
  painter.release
end

image.save("preview.png")
\end{lstlisting}

The same rule applies to \api{QPixmap}, \api{QSvgGenerator}, and \api{QPdfWriter}. Finish painting first; export or inspect second.

\section{\api{release}}

Both \api{QObject} and \api{NativeResource} implement \api{release}. Calling \api{release} asks the wrapper to destroy its native object, but only if the wrapper owns it. If the object is parent-owned or already destroyed, \api{release} is a no-op.

\begin{lstlisting}[style=crystal]
dialog = Qt6::InputDialog.new(main)
dialog.window_title = "Rename Layer"

begin
  if dialog.exec == Qt6::DialogCode::Accepted
    puts dialog.text_value
  end
ensure
  dialog.release
end
\end{lstlisting}

The convenience dialog helpers use this pattern internally. They create a dialog, run it modally, return the chosen value, and release the native dialog before returning.

\begin{lstlisting}[style=crystal]
name = Qt6::InputDialog.get_text(
  main,
  title: "Rename Layer",
  label: "Layer name",
  value: "Terrain"
)
\end{lstlisting}

For child widgets and parented objects, prefer parent ownership over manual release. Manual release is best for short-lived unparented resources whose lifetime is obvious.

\section{Release Checklist}

Use this checklist before calling \api{release}:

\begin{enumerate}
\item Did you create the object without a parent, and does the wrapper own it?
\item Is the object absent from any parent widget, layout, menu, toolbar, model, or stack that still expects to use it?
\item Is the lifetime short and local enough that \api{ensure} can release it reliably?
\item For paint devices, has the active \api{QPainter} been released before the device is saved or read?
\item For large native resources, is early cleanup useful enough to make manual lifetime management clearer rather than noisier?
\end{enumerate}

Prefer parent ownership when the object belongs to the UI. Prefer automatic shutdown for ordinary top-level examples. Prefer explicit \api{release} for short-lived dialogs, temporary painters, large transient images, and other unparented native resources with a clear end point.

\section{Tracked Shutdown}

The first call to \api{Qt6.application} registers an \api{at_exit} hook. At shutdown, \api{Qt6.shutdown} releases tracked objects in reverse creation order and then destroys the shared application wrapper.

\begin{lstlisting}[style=crystal]
app = Qt6.application
window = Qt6.window("Example", 320, 160) do |widget|
  widget.vbox do |layout|
    layout << Qt6::Label.new("Hello")
  end
end

window.show
app.run
# Qt6.shutdown runs automatically at process exit.
\end{lstlisting}

You can call \api{Qt6.shutdown} yourself if your program has a clear boundary where all Qt work is finished. Most GUI applications do not need to. The important part is that Qt objects are not left to Crystal's garbage collector to destroy at arbitrary times.

\section{Shutdown Order For Contributors}

Tracked shutdown releases wrapper-owned objects in reverse creation order. That detail is mainly for contributors and tests, not day-to-day application code. Reverse order helps with the common case where a later-created helper depends on an earlier-created application object or window.

Do not design application logic around the exact shutdown order. If one object must die before another, express that relationship with Qt parent ownership, a local \api{ensure} block, or an explicit application-level teardown method.

\section{Borrowed Handles And Wrapped Objects}

Some methods return wrappers around native objects owned elsewhere. These wrappers are borrowed views of Qt state. For example, \api{main.status_bar} returns the main window's status bar, and \api{Qt6.clipboard} returns the process-wide clipboard wrapper.

\begin{lstlisting}[style=crystal]
status = main.status_bar
status.show_message("Ready")

clipboard = Qt6.clipboard
clipboard.text = "Copied from Crystal"
\end{lstlisting}

Borrowed wrappers should not be manually released. Their native lifetime belongs to the object that returned them. Use them while the owner is alive.

\section{Signals And Callback Lifetime}

Signal helper methods such as \api{on_clicked}, \api{on_triggered}, and \api{on_timeout} store Crystal callbacks on the wrapper. The native Qt signal routes back through that wrapper.

\begin{lstlisting}[style=crystal]
timer = Qt6::QTimer.new(main)
timer.on_timeout do
  main.status_bar.show_message("Autosaved", 1500)
end
timer.start(5_000)
\end{lstlisting}

This is another reason to keep ordinary GUI objects parented to a durable owner. If the timer belongs to the main window, the timer's native lifetime and its Crystal callback lifetime both naturally end with that window.

\section{GUI Thread Expectations}

Qt Widgets are GUI-thread objects. Create widgets, show windows, mutate widget state, and destroy widget wrappers from the same thread that owns the application event loop. This is the safest rule even for operations that look small, such as changing a label or releasing a dialog.

Use background work only for ordinary Crystal computation or blocking I/O that does not touch Qt objects directly. Move the result back to the GUI flow before updating widgets, models, images used by widgets, or actions. If the binding later grows explicit thread or queued-callback helpers, those helpers should become the boundary between worker code and GUI code.

Timers, signal callbacks, paint callbacks, and event callbacks already run in the Qt event flow. Keep them brief, update application state, schedule repaint with \api{update}, and return control to the event loop.

\section{Common Pitfalls}

The common ownership mistakes are predictable:

\begin{enumerate}
\item Releasing a parent-owned object manually, then expecting the parent to keep using it.
\item Keeping a wrapper in a long-lived variable after its native owner has been destroyed.
\item Creating many unparented objects in a long-running workflow and relying entirely on process shutdown.
\item Treating borrowed wrappers, such as a status bar or clipboard wrapper, as objects you own.
\item Creating or destroying widgets from background threads instead of the GUI thread.
\end{enumerate}

The usual remedy is to make ownership boring. Give child objects a parent, keep top-level windows alive for as long as the UI needs them, use convenience helpers for modal dialogs, and reserve explicit \api{release} for short-lived unparented resources.

\section{Practical Guidance}

For application code, these defaults work well:

\begin{enumerate}
\item Create one application object with \api{Qt6.application}.
\item Parent controls, timers, actions, models, and dialogs to a durable widget or object.
\item Let top-level windows be wrapper-owned.
\item Use \api{app.quit} to leave the event loop.
\item Let \api{Qt6.shutdown} run automatically unless you have a clear reason to call it yourself.
\end{enumerate}

That model is deliberately conservative. It matches Qt's expectations, keeps Crystal's garbage collector out of native teardown decisions, and gives larger applications a stable base for menus, dialogs, custom widgets, rendering resources, and model/view objects.
