\chapter{Widgets And Layouts}

Widgets are the visible building blocks of a Qt application. Layouts decide how those widgets share space. This chapter introduces the core widget surface in \api{crystal-qt6}: common controls, sizing, attributes, layout composition, styling, and small bits of top-level window polish.

\section{The Widget Base}

\api{Qt6::Widget} wraps \api{QWidget}, the base class for most visible controls. It exposes shared behavior such as titles, visibility, enabled state, focus, tooltips, sizing, style sheets, cursor shape, drop acceptance, widget attributes, repaint scheduling, and screenshot capture.

\begin{lstlisting}[style=crystal]
panel = Qt6::Widget.new
panel.tool_tip = "Layer inspector"
panel.enabled = true
panel.set_minimum_size(280, 220)
panel.set_size_policy(Qt6::SizePolicy::Preferred, Qt6::SizePolicy::Expanding)
\end{lstlisting}

A widget can be top-level, parented to another widget, or placed inside a layout. Top-level widgets need a title and usually an initial size. Child widgets usually let their layout choose the final geometry.

\begin{lstlisting}[style=crystal]
window = Qt6::Widget.new
window.window_title = "Inspector"
window.resize(420, 320)
window.show
\end{lstlisting}

\section{Core Controls}

The simplest controls are labels and buttons. Labels show text. Buttons emit callbacks when clicked.

\begin{lstlisting}[style=crystal]
count = 0
value = Qt6::Label.new("0")
button = Qt6::PushButton.new("Increment")

button.on_clicked do
  count += 1
  value.text = count.to_s
end
\end{lstlisting}

Forms are usually built from a small family of input widgets:

\begin{lstlisting}[style=crystal]
name = Qt6::LineEdit.new("Terrain")
visible = Qt6::CheckBox.new("Visible")
kind = Qt6::ComboBox.new
kind << "Hexes" << "Terrain" << "Units"

opacity = Qt6::SpinBox.new
opacity.set_range(0, 100)
opacity.value = 72

scale = Qt6::DoubleSpinBox.new
scale.set_range(0.25, 4.0)
scale.value = 1.0
\end{lstlisting}

The \api{crystal-qt6} library also includes radio buttons, sliders, group boxes, progress bars, text editors, item widgets, date/time controls, tab widgets, scroll areas, splitters, and other common desktop controls. Later chapters use those in larger shell and editor examples.

\section{Visibility, Enabled State, And Focus}

Visibility and enabled state are shared widget features:

\begin{lstlisting}[style=crystal]
advanced_panel.visible = show_advanced.checked?
apply_button.enabled = !name.text.empty?
\end{lstlisting}

Focus is explicit when you need keyboard input to land in a particular control.

\begin{lstlisting}[style=crystal]
name.focus_policy = Qt6::FocusPolicy::StrongFocus
name.set_focus
\end{lstlisting}

For ordinary forms, Qt's default focus behavior is usually enough. Use explicit focus control for editor workflows, search fields, modal flows, and custom canvases.

\section{Sizing}

Qt layouts negotiate size from several pieces of information: each widget's size hint, its minimum and maximum size, and its size policy. \api{crystal-qt6} exposes the common controls for that negotiation.

\begin{lstlisting}[style=crystal]
preview = Qt6::Label.new("Preview")
preview.set_minimum_size(240, 160)
preview.maximum_height = 240

swatch = Qt6::Label.new("")
swatch.set_fixed_size(96, 24)

notes = Qt6::PlainTextEdit.new
notes.set_size_policy(Qt6::SizePolicy::Expanding, Qt6::SizePolicy::Expanding)
\end{lstlisting}

Use fixed sizes sparingly. They are appropriate for swatches, icon previews, compact counters, and controls whose physical dimensions are part of the UI. For panels and editors, prefer minimum sizes plus expanding or preferred size policies.

\section{Widget Attributes And Cursors}

Qt widget attributes are lower-level flags for behavior that is not always worth a dedicated high-level method. \api{crystal-qt6} exposes them through \api{WidgetAttribute}.

\begin{lstlisting}[style=crystal]
overlay = Qt6::Label.new("Drop files here")
overlay.set_attribute(Qt6::WidgetAttribute::StyledBackground)
overlay.set_attribute(Qt6::WidgetAttribute::TransparentForMouseEvents)
\end{lstlisting}

Some attributes also have convenience methods. Prefer the clearer method when one exists:

\begin{lstlisting}[style=crystal]
canvas = Qt6::EventWidget.new
canvas.mouse_tracking = true
canvas.accept_drops = true
canvas.cursor_shape = Qt6::CursorShape::Cross
\end{lstlisting}

Attributes are powerful, but they are close to Qt itself. Use them to express known Qt behavior, not as a substitute for ordinary widget APIs.

\section{Layout Families}

Qt layouts are parent-owned objects that place child widgets. In Crystal, \api{Widget#vbox}, \api{#hbox}, \api{#form}, and \api{#grid} create the corresponding layout and yield it for composition.

\api{VBoxLayout} stacks widgets vertically:

\begin{lstlisting}[style=crystal]
panel = Qt6::Widget.new
panel.vbox do |column|
  column << Qt6::Label.new("Layers")
  column << layer_list
  column << apply_button
end
\end{lstlisting}

\api{HBoxLayout} is useful for command rows:

\begin{lstlisting}[style=crystal]
buttons = Qt6::Widget.new
buttons.hbox do |row|
  row << Qt6::PushButton.new("Add")
  row << Qt6::PushButton.new("Remove")
  row << Qt6::PushButton.new("Apply")
end
\end{lstlisting}

\api{FormLayout} is best for labeled editor fields:

\begin{lstlisting}[style=crystal]
properties = Qt6::Widget.new
properties.form do |form|
  form.add_row("Name", name)
  form.add_row("Kind", kind)
  form.add_row("Opacity", opacity)
  form.add_row(visible)
end
\end{lstlisting}

\api{GridLayout} works well for matrix-like tools, command palettes, and compact dashboards:

\begin{lstlisting}[style=crystal]
grid = Qt6::Widget.new
grid.grid do |layout|
  layout.add(Qt6::Label.new("X"), 0, 0)
  layout.add(Qt6::Label.new("Y"), 0, 1)
  layout.add(x_spin, 1, 0)
  layout.add(y_spin, 1, 1)
end
\end{lstlisting}

\section{Composing A Panel}

Most desktop panels combine several layouts. A common pattern is a vertical column containing a form plus a row of actions.

\begin{lstlisting}[style=crystal]
name = Qt6::LineEdit.new("Terrain")
kind = Qt6::ComboBox.new
kind << "Hexes" << "Terrain" << "Units"

visible = Qt6::CheckBox.new("Visible")
apply = Qt6::PushButton.new("Apply")
reset = Qt6::PushButton.new("Reset")

panel = Qt6::Widget.new
panel.set_minimum_size(320, 220)
panel.vbox do |column|
  form_host = Qt6::Widget.new
  form_host.form do |form|
    form.add_row("Name", name)
    form.add_row("Kind", kind)
    form.add_row(visible)
  end

  actions = Qt6::Widget.new
  actions.hbox do |row|
    row << reset
    row << apply
  end

  column << form_host
  column << actions
end
\end{lstlisting}

This is not the only way to compose a panel, but it scales well. The form stays responsible for labels and fields. The horizontal layout stays responsible for commands. The outer vertical layout controls overall flow.

\section{Styling}

Qt style sheets are CSS-like strings applied to widgets or the whole application. They are useful for lightweight polish, warnings, previews, and demo styling.

\begin{lstlisting}[style=crystal]
swatch = Qt6::Label.new("")
swatch.set_fixed_size(96, 24)
swatch.style_sheet = "background-color: #3470b0; border: 1px solid #6b7280;"

name.tool_tip = "Layer name shown in the layer list"
\end{lstlisting}

Application-wide style sheets are available from the application wrapper:

\begin{lstlisting}[style=crystal]
app = Qt6.application
app.style_sheet = "
  QGroupBox { font-weight: bold; }
  QPushButton { padding: 4px 10px; }
"
\end{lstlisting}

Use style sheets to support the interface rather than replace layout. Sizing, hierarchy, and interaction should come from the widget tree and layout structure first.

\section{Window Polish}

Top-level widgets and windows expose title, icon, size, visibility, and capture helpers.

\begin{lstlisting}[style=crystal]
main = Qt6::MainWindow.new
main.window_title = "Map Tool"
main.resize(960, 640)

icon = Qt6::QIcon.new("assets/app-icon.png")
main.window_icon = icon

main.status_bar.show_message("Ready")
main.show
\end{lstlisting}

The application object can also hold application-wide metadata and window icon state:

\begin{lstlisting}[style=crystal]
app = Qt6.application
app.name = "Map Tool"
app.organization_name = "Example Studio"
app.window_icon = icon
\end{lstlisting}

For quick exports and documentation screenshots, \api{Widget#grab} captures a widget into a \api{QPixmap}.

\begin{lstlisting}[style=crystal]
pixmap = main.grab
pixmap.save("window.png")
\end{lstlisting}

\section{Where To Look Next}

\api{examples/inspector_workbench.cr} is the best example for form controls, group boxes, sliders, spin boxes, tabs, splitters, and a live custom canvas. \api{examples/editor_shell.cr} shows how ordinary widgets fit into a main-window shell with actions, menus, toolbars, docks, dialogs, and a status bar.
