\chapter{Undo And Redo}

Undo support is built around \api{UndoStack} and Crystal-backed \api{UndoCommand}. A command describes one reversible edit. The stack owns the history, tracks the current index, exposes clean/dirty state, and creates ordinary \api{Action} objects for menus and toolbars.

Qt uses a simple rule: pushing a command calls its \api{redo} behavior immediately. That means an undo command is not just a historical record; it is the object that performs the edit, reverses it, and performs it again.

\section{The Command Model}

A command needs a user-facing label and two callbacks: redo and undo. The label appears in action text such as \api{Undo Rename layer}.

\begin{lstlisting}[style=crystal]
stack = Qt6::UndoStack.new(main)
layer_name = "Terrain"

command = Qt6::UndoCommand.new(
  "Rename layer",
  redo: -> { layer_name = "Roads" },
  undo: -> { layer_name = "Terrain" }
)

stack.push(command) # calls redo, so layer_name is now "Roads"
\end{lstlisting}

The same command can be written with setters when that reads better:

\begin{lstlisting}[style=crystal]
command = Qt6::UndoCommand.new("Rename layer")
command.on_redo { layer_name = "Roads" }
command.on_undo { layer_name = "Terrain" }

stack.push(command)
\end{lstlisting}

After a command is pushed, the native stack owns it. Do not call \api{release} on a pushed command. The Crystal wrapper keeps the callbacks alive until Qt destroys the command.

\section{Capture Before And After State}

Most commands should capture the old value and the new value before they are pushed. The redo callback applies the new value; the undo callback restores the old value.

\begin{lstlisting}[style=crystal]
old_name = layer.name
new_name = name_edit.text

if old_name != new_name
  stack.push(Qt6::UndoCommand.new(
    "Rename layer",
    redo: -> {
      layer.name = new_name
      name_label.text = new_name
      canvas.update
    },
    undo: -> {
      layer.name = old_name
      name_label.text = old_name
      canvas.update
    }
  ))
end
\end{lstlisting}

Notice that both directions update the model and the view. A command should leave the application in a complete state whether it is run forward or backward.

\section{Undo And Redo Actions}

\api{UndoStack} can create \api{Action} objects that are already connected to the stack. Add those actions to menus and toolbars the same way you would add any other application command.

\begin{lstlisting}[style=crystal]
undo_action = stack.create_undo_action(main, "Undo")
undo_action.shortcut = "Ctrl+Z"

redo_action = stack.create_redo_action(main, "Redo")
redo_action.shortcut = "Ctrl+Shift+Z"

edit_menu = main.menu_bar.add_menu("Edit")
edit_menu << undo_action
edit_menu << redo_action

toolbar << undo_action
toolbar << redo_action
\end{lstlisting}

The actions update their enabled state and text as the stack changes. When the next undo step is \api{Rename layer}, the undo action can present that operation to the user.

\section{Stack State}

The stack exposes the current history position and user-facing text for the next undo or redo step.

\begin{longtable}{@{}p{0.30\textwidth}p{0.64\textwidth}@{}}
\toprule
API & Meaning \\
\midrule
\api{can_undo?} & Returns whether an undo step is available. \\
\api{can_redo?} & Returns whether a redo step is available. \\
\api{undo_text} & Returns the label for the next undo step. \\
\api{redo_text} & Returns the label for the next redo step. \\
\api{count} & Returns the number of retained history entries. \\
\api{index} & Returns the current position in the history. \\
\api{clean_index} & Returns the saved-document history position, or \api{-1} when there is no clean marker. \\
\api{undo_limit} & Limits retained commands; set it before pushing commands. \\
\api{clear} & Removes all commands from the stack. \\
\bottomrule
\end{longtable}

\begin{lstlisting}[style=crystal]
status_bar.show_message(
  "#{stack.index}/#{stack.count} undo: #{stack.undo_text}",
  1500
)

stack.undo if stack.can_undo?
stack.redo if stack.can_redo?
\end{lstlisting}

A stack index is a cursor between commands. If the user undoes a few commands and then pushes a new command, Qt discards the redo tail and appends the new command at the current index.

\begin{figure}[htbp]
  \centering
  \scriptsize
  \setlength{\tabcolsep}{3pt}
  \begin{tabular}{@{}p{0.16\textwidth}p{0.16\textwidth}p{0.16\textwidth}p{0.16\textwidth}p{0.16\textwidth}@{}}
  \toprule
  State & Command 1 & Command 2 & Command 3 & Command 4 \\
  \midrule
  Before new push & Rename layer & Move point & Recolor shape & Resize shape \\
  Markers & & clean + current index & redo tail & redo tail \\
  \midrule
  After pushing & Rename layer & Move point & Set opacity & discarded \\
  Markers & & clean index & current index & removed redo tail \\
  \bottomrule
  \end{tabular}
  \caption{Undo history after the user undoes to the clean index, then pushes a new command. The old redo branch is discarded.}
\end{figure}

\section{Clean State}

The clean marker records which stack index corresponds to the saved document. When the current index equals the clean index, \api{clean?} is true.

\begin{lstlisting}[style=crystal]
save_action.on_triggered do
  save_document
  stack.set_clean
end

stack.on_clean_changed do |clean|
  save_action.enabled = !clean
  main.window_title = clean ? "Map Tool" : "Map Tool *"
end
\end{lstlisting}

Use \api{set_clean} after a successful save. Use \api{reset_clean} when there is no meaningful saved state anymore, such as after loading a new unsaved document or importing content that cannot be matched to a file.

\begin{lstlisting}[style=crystal]
stack.clear
load_document(path)
stack.set_clean
\end{lstlisting}

If the user undoes back to the saved index, \api{clean?} becomes true again. That is one reason clean state is more useful than a single boolean dirty flag.

\section{Macros}

Use macros to group several commands into one user-visible history entry. This is appropriate when one user gesture changes several pieces of state.

\begin{lstlisting}[style=crystal]
stack.begin_macro("Apply layer preset")

stack.push(Qt6::UndoCommand.new(
  "Set color",
  redo: -> { layer.color = preset.color },
  undo: -> { layer.color = old_color }
))

stack.push(Qt6::UndoCommand.new(
  "Set opacity",
  redo: -> { layer.opacity = preset.opacity },
  undo: -> { layer.opacity = old_opacity }
))

stack.end_macro
\end{lstlisting}

The stack will show one entry, \api{Apply layer preset}. Undoing that entry runs the child commands backward; redoing it runs them forward.

Always balance \api{begin_macro} with \api{end_macro}. If the operation can fail in the middle, gather the values first, then open the macro only when you are ready to push all of its commands.

\begin{lstlisting}[style=crystal]
def with_undo_macro(stack : Qt6::UndoStack, text : String, &block)
  stack.begin_macro(text)
  begin
    yield
  ensure
    stack.end_macro
  end
end

with_undo_macro(stack, "Apply layer preset") do
  stack.push(set_color_command)
  stack.push(set_opacity_command)
end
\end{lstlisting}

The helper does not make the operation transactional; commands pushed before an exception are still part of the macro. Its job is narrower but important: keep the native stack balanced so later undo operations remain predictable.

\section{Stack Signals}

\api{UndoStack} emits signals whenever availability, clean state, index, or action text changes.

\begin{lstlisting}[style=crystal]
stack.on_can_undo_changed do |available|
  undo_badge.visible = available
end

stack.on_index_changed do |index|
  history_label.text = "#{index} / #{stack.count}"
end

stack.on_undo_text_changed do |text|
  status_bar.show_message(text.empty? ? "Nothing to undo" : "Next undo: #{text}", 1200)
end
\end{lstlisting}

The generated undo and redo actions already listen to the stack. Use these signals for secondary UI: title bars, status bars, history labels, dirty indicators, or custom command palettes.

\section{Text Editing}

Text widgets have their own built-in undo history. Use that for ordinary text editing inside \api{TextEdit}, \api{PlainTextEdit}, and \api{TextDocument}.

\begin{lstlisting}[style=crystal]
notes = Qt6::PlainTextEdit.new
notes.undo_redo_enabled = true

notes.insert_plain_text("Roads")
notes.undo if notes.can_undo?
notes.redo if notes.can_redo?
\end{lstlisting}

Use an application-level \api{UndoStack} for domain edits: renaming layers, changing map objects, importing assets, moving nodes, applying presets, or changing application modes. Do not mirror every keystroke from a text editor into your application stack unless the text edit itself is the document model you are building.

\section{A Small Pattern}

It is useful to hide repetitive command setup behind small helper functions. This keeps UI handlers readable and makes all undoable edits follow the same shape.

\begin{lstlisting}[style=crystal]
def push_change(stack, text, apply, revert)
  stack.push(Qt6::UndoCommand.new(text, redo: apply, undo: revert))
end

old_visible = layer.visible?
new_visible = visible_check.checked?

push_change(
  stack,
  new_visible ? "Show layer" : "Hide layer",
  -> {
    layer.visible = new_visible
    canvas.update
  },
  -> {
    layer.visible = old_visible
    canvas.update
  }
)
\end{lstlisting}

The command owns the edit. The button, checkbox, or menu action simply decides which command to push.

\section{A Small Layer Example}

The following mini-application keeps a list of layer names, exposes undo and redo through the Edit menu, marks the document clean after a save-like action, and groups a preset operation into one visible history entry.

\begin{lstlisting}[style=crystal]
require "qt6"

def with_undo_macro(stack : Qt6::UndoStack, text : String, &block)
  stack.begin_macro(text)
  begin
    yield
  ensure
    stack.end_macro
  end
end

app = Qt6.application

main = Qt6::MainWindow.new
main.window_title = "Layer Undo"
main.resize(420, 300)

stack = Qt6::UndoStack.new(main)
layers = ["Terrain", "Roads"]

central = Qt6::Widget.new
main.central_widget = central

list = Qt6::ListWidget.new
status = Qt6::Label.new
add_button = Qt6::PushButton.new("Add layer")
rename_button = Qt6::PushButton.new("Rename first")
preset_button = Qt6::PushButton.new("Apply preset")

refresh = -> do
  list.clear
  layers.each { |name| list << name }
  status.text = stack.clean? ? "Saved" : "Unsaved changes"
  main.window_title = stack.clean? ? "Layer Undo" : "Layer Undo *"
end

central.vbox do |column|
  column << list
  column << status
  column << add_button
  column << rename_button
  column << preset_button
end

undo_action = stack.create_undo_action(main, "Undo")
undo_action.shortcut = "Ctrl+Z"
redo_action = stack.create_redo_action(main, "Redo")
redo_action.shortcut = "Ctrl+Shift+Z"

save_action = Qt6::Action.new("Mark Saved", main)
save_action.shortcut = "Ctrl+S"
save_action.on_triggered do
  stack.set_clean
  refresh.call
end

edit_menu = main.menu_bar.add_menu("Edit")
edit_menu << undo_action
edit_menu << redo_action
edit_menu.add_separator
edit_menu << save_action

stack.on_clean_changed { |_clean| refresh.call }
stack.on_index_changed { |_index| refresh.call }

add_button.on_clicked do
  name = "Layer #{layers.size + 1}"
  stack.push(Qt6::UndoCommand.new(
    "Add #{name}",
    redo: -> { layers << name; refresh.call },
    undo: -> { layers.delete(name); refresh.call }
  ))
end

rename_button.on_clicked do
  next if layers.empty?

  old_name = layers.first
  new_name = "#{old_name} copy"
  stack.push(Qt6::UndoCommand.new(
    "Rename layer",
    redo: -> { layers[0] = new_name; refresh.call },
    undo: -> { layers[0] = old_name; refresh.call }
  ))
end

preset_button.on_clicked do
  next if layers.empty?

  old_name = layers.first
  added_name = "Highlights"

  with_undo_macro(stack, "Apply starter preset") do
    stack.push(Qt6::UndoCommand.new(
      "Rename base layer",
      redo: -> { layers[0] = "Background"; refresh.call },
      undo: -> { layers[0] = old_name; refresh.call }
    ))
    stack.push(Qt6::UndoCommand.new(
      "Add highlights",
      redo: -> { layers << added_name; refresh.call },
      undo: -> { layers.delete(added_name); refresh.call }
    ))
  end
end

refresh.call
stack.set_clean

main.show
app.run
\end{lstlisting}

\section{Command Compression}

Some applications should combine many tiny edits into one undoable command: dragging an object, scrubbing a slider, or typing into a domain-specific property editor. Qt's native command model supports this through command IDs and merge behavior, but those APIs are not exposed by the binding yet.

Until merge APIs are available, coalesce changes before pushing. For example, push one command when a drag finishes rather than on every mouse move, or push one command when a field commits rather than on every keystroke. Use macros when one completed gesture changes several properties at once.

\section{Multiple Documents}

Use one \api{UndoStack} per document. Clear or replace that stack when the document changes identity, and mark it clean only after the corresponding document is saved.

\api{UndoGroup} coordinates a set of document stacks and creates shared undo and redo actions that follow whichever stack is active. This is the usual shape for tabbed editors, multiple document interfaces, or workbenches where the Edit menu should stay in one place while focus moves between documents.

\begin{lstlisting}[style=crystal]
undo_group = Qt6::UndoGroup.new(main)
undo_group << terrain_document.undo_stack
undo_group << roads_document.undo_stack

undo_action = undo_group.create_undo_action(main, "Undo")
redo_action = undo_group.create_redo_action(main, "Redo")
edit_menu << undo_action
edit_menu << redo_action

tabs.on_current_changed do |index|
  undo_group.active_stack = documents[index].undo_stack
end
\end{lstlisting}

The group does not replace document ownership. Each document still owns its own \api{UndoStack}; the group only decides which stack receives the shared undo and redo commands.

\section{Practical Guidelines}

Name commands from the user's point of view: \api{Rename layer}, \api{Move waypoint}, \api{Delete selection}, \api{Apply preset}. Short labels make undo and redo actions feel natural.

Do not push commands for no-op changes. If the old value and new value are equal, skip the command.

Make redo and undo symmetrical. Each direction should update model state, derived state, and visible UI that depends on the edit.

Set \api{undo_limit} before commands are pushed if the document can generate a large history. Clear the stack when the document changes identity.

Prefer macros for one user gesture that touches several fields. Prefer separate commands for separate gestures, even when they happen close together in time.
