\chapter{Coverage and Reference Appendix}

This appendix summarizes the current \api{crystal-qt6} API shape at a high level. Treat the README and specs as the executable source of truth when release boundaries move.

\begin{longtable}{@{}p{0.26\textwidth}p{0.66\textwidth}@{}}
\toprule
Area & Representative APIs \\
\midrule
Application & \api{Application}, \api{Qt6.application}, metadata, style sheets, clipboard, local event loops \\
Shells & \api{MainWindow}, \api{Action}, \api{ActionGroup}, \api{MenuBar}, \api{Menu}, \api{ToolBar}, \api{StatusBar}, \api{DockWidget} \\
Dialogs & \api{Dialog}, \api{MessageBox}, \api{FileDialog}, \api{ColorDialog}, \api{FontDialog}, \api{InputDialog}, \api{ProgressDialog}, \api{SplashScreen}, \api{DialogButtonBox} \\
Widgets & Labels, buttons, line edits, check boxes, radio buttons, combo boxes, sliders, spin boxes, tabs, splitters, scroll areas, group boxes, progress bars, date/time controls, calendar widgets, text editors \\
Events & \api{EventWidget}, paint, resize, mouse, wheel, key, focus, enter/leave, drag/drop callbacks, synthetic event helpers \\
Rendering & \api{QPainter}, \api{QPainterPath}, \api{QTransform}, \api{QPen}, \api{QBrush}, \api{QFont}, font metrics, SVG, PDF \\
Images & \api{QImage}, \api{QPixmap}, \api{QImageReader}, raw buffers, pixel operations, conversion, transforms, scaling, clipboard images \\
Undo & \api{UndoStack}, \api{UndoCommand}, clean state, macros, undo/redo actions, stack signals \\
Model/View & Item widgets, abstract list/tree models, standard item models, proxy sorting/filtering, selection models, delegates, list/tree/table views \\
\bottomrule
\end{longtable}

\section{Enum Reference}

The following tables collect enum values that appear often in examples and callbacks. They are a reader-facing snapshot of the Crystal wrappers, not a substitute for checking the source when adding new bindings.

Flags enums can be combined with Crystal's bitwise OR operator. For example, an item can be made selectable and enabled with \api{Qt6::ItemFlag::Selectable | Qt6::ItemFlag::Enabled}.

\subsection{Event Types}

\api{EventType} values mirror Qt's event type numbers. Most applications handle these indirectly through higher-level callbacks, but the numeric values are useful when inspecting generic events, logs, or test helpers.

{\small
\begin{longtable}{@{}p{0.38\textwidth}p{0.10\textwidth}p{0.44\textwidth}@{}}
\toprule
Name & Value & Typical use \\
\midrule
\api{None} & 0 & No event or an unspecified event. \\
\api{Timer} & 1 & Timer delivery. \\
\api{MouseButtonPress} & 2 & Mouse or pointer button press. \\
\api{MouseButtonRelease} & 3 & Mouse or pointer button release. \\
\api{MouseButtonDblClick} & 4 & Mouse double-click. \\
\api{MouseMove} & 5 & Mouse or pointer movement. \\
\api{KeyPress} & 6 & Keyboard press. \\
\api{KeyRelease} & 7 & Keyboard release. \\
\api{FocusIn} & 8 & Widget receives keyboard focus. \\
\api{FocusOut} & 9 & Widget loses keyboard focus. \\
\api{Enter} & 10 & Pointer enters a widget. \\
\api{Leave} & 11 & Pointer leaves a widget. \\
\api{Paint} & 12 & Widget repaint request. \\
\api{Move} & 13 & Widget position changes. \\
\api{Resize} & 14 & Widget size changes. \\
\api{Create} & 15 & Native widget creation. \\
\api{Destroy} & 16 & Native widget destruction. \\
\api{Show} & 17 & Widget becomes visible. \\
\api{Hide} & 18 & Widget becomes hidden. \\
\api{Close} & 19 & Window close request. \\
\api{Quit} & 20 & Application quit request. \\
\api{ParentChange} & 21 & Object or widget parent changes. \\
\api{ThreadChange} & 22 & Object affinity moves to another thread. \\
\api{FocusAboutToChange} & 23 & Focus is about to move. \\
\api{WindowActivate} & 24 & Window becomes active. \\
\api{WindowDeactivate} & 25 & Window becomes inactive. \\
\api{ShowToParent} & 26 & Child widget is shown to its parent. \\
\api{HideToParent} & 27 & Child widget is hidden from its parent. \\
\api{Wheel} & 31 & Mouse wheel or trackpad scroll. \\
\api{WindowTitleChange} & 33 & Window title changes. \\
\api{WindowIconChange} & 34 & Window icon changes. \\
\api{ApplicationWindowIconChange} & 35 & Application window icon changes. \\
\api{ApplicationFontChange} & 36 & Application font changes. \\
\api{ApplicationLayoutDirectionChange} & 37 & Application layout direction changes. \\
\api{ApplicationPaletteChange} & 38 & Application palette changes. \\
\api{PaletteChange} & 39 & Widget palette changes. \\
\api{Clipboard} & 40 & Clipboard contents change. \\
\api{Speech} & 42 & Platform speech event. \\
\api{MetaCall} & 43 & Queued signal, slot, or meta-call delivery. \\
\api{SockAct} & 50 & Socket notifier activity. \\
\api{ShortcutOverride} & 51 & Shortcut override opportunity before key handling. \\
\api{DeferredDelete} & 52 & Deferred object deletion. \\
\api{DragEnter} & 60 & Drag operation enters a widget. \\
\api{DragMove} & 61 & Drag operation moves over a widget. \\
\api{DragLeave} & 62 & Drag operation leaves a widget. \\
\api{Drop} & 63 & Drop operation completes. \\
\bottomrule
\end{longtable}
}

\subsection{Common UI Enums}

{\small
\begin{longtable}{@{}p{0.34\textwidth}p{0.42\textwidth}p{0.12\textwidth}@{}}
\toprule
Enum & Name & Value \\
\midrule
\api{DialogCode} & \api{Rejected} & 0 \\
\api{DialogCode} & \api{Accepted} & 1 \\
\api{Orientation} & \api{Horizontal} & 1 \\
\api{Orientation} & \api{Vertical} & 2 \\
\api{SortOrder} & \api{Ascending} & 0 \\
\api{SortOrder} & \api{Descending} & 1 \\
\api{CheckState} & \api{Unchecked} & 0 \\
\api{CheckState} & \api{PartiallyChecked} & 1 \\
\api{CheckState} & \api{Checked} & 2 \\
\api{FileDialogAcceptMode} & \api{Open} & 0 \\
\api{FileDialogAcceptMode} & \api{Save} & 1 \\
\api{FileDialogFileMode} & \api{AnyFile} & 0 \\
\api{FileDialogFileMode} & \api{ExistingFile} & 1 \\
\api{FileDialogFileMode} & \api{Directory} & 2 \\
\api{FileDialogFileMode} & \api{ExistingFiles} & 3 \\
\bottomrule
\end{longtable}
}

\subsection{Model/View Enums and Flags}

{\small
\begin{longtable}{@{}p{0.34\textwidth}p{0.42\textwidth}p{0.12\textwidth}@{}}
\toprule
Enum & Name & Value \\
\midrule
\api{ItemDataRole} & \api{Display} & 0 \\
\api{ItemDataRole} & \api{Decoration} & 1 \\
\api{ItemDataRole} & \api{Edit} & 2 \\
\api{ItemDataRole} & \api{ToolTip} & 3 \\
\api{ItemDataRole} & \api{StatusTip} & 4 \\
\api{ItemDataRole} & \api{WhatsThis} & 5 \\
\api{ItemDataRole} & \api{Font} & 6 \\
\api{ItemDataRole} & \api{TextAlignment} & 7 \\
\api{ItemDataRole} & \api{Background} & 8 \\
\api{ItemDataRole} & \api{Foreground} & 9 \\
\api{ItemDataRole} & \api{CheckState} & 10 \\
\api{ItemDataRole} & \api{AccessibleText} & 11 \\
\api{ItemDataRole} & \api{AccessibleDescription} & 12 \\
\api{ItemDataRole} & \api{SizeHint} & 13 \\
\api{ItemDataRole} & \api{InitialSortOrder} & 14 \\
\api{ItemDataRole} & \api{User} & 256 \\
\api{ItemFlag} & \api{None} & 0 \\
\api{ItemFlag} & \api{Selectable} & 1 \\
\api{ItemFlag} & \api{Editable} & 2 \\
\api{ItemFlag} & \api{DragEnabled} & 4 \\
\api{ItemFlag} & \api{DropEnabled} & 8 \\
\api{ItemFlag} & \api{UserCheckable} & 16 \\
\api{ItemFlag} & \api{Enabled} & 32 \\
\api{ItemFlag} & \api{AutoTristate} & 64 \\
\api{ItemFlag} & \api{NeverHasChildren} & 128 \\
\api{ItemFlag} & \api{UserTristate} & 256 \\
\api{ItemSelectionMode} & \api{NoSelection} & 0 \\
\api{ItemSelectionMode} & \api{SingleSelection} & 1 \\
\api{ItemSelectionMode} & \api{MultiSelection} & 2 \\
\api{ItemSelectionMode} & \api{ExtendedSelection} & 3 \\
\api{ItemSelectionMode} & \api{ContiguousSelection} & 4 \\
\api{ItemSelectionBehavior} & \api{SelectItems} & 0 \\
\api{ItemSelectionBehavior} & \api{SelectRows} & 1 \\
\api{ItemSelectionBehavior} & \api{SelectColumns} & 2 \\
\api{SelectionFlag} & \api{NoUpdate} & 0 \\
\api{SelectionFlag} & \api{Clear} & 1 \\
\api{SelectionFlag} & \api{Select} & 2 \\
\api{SelectionFlag} & \api{Deselect} & 4 \\
\api{SelectionFlag} & \api{Toggle} & 8 \\
\api{SelectionFlag} & \api{Current} & 16 \\
\api{SelectionFlag} & \api{Rows} & 32 \\
\api{SelectionFlag} & \api{Columns} & 64 \\
\api{SelectionFlag} & \api{SelectCurrent} & 18 \\
\api{SelectionFlag} & \api{ToggleCurrent} & 24 \\
\api{SelectionFlag} & \api{ClearAndSelect} & 3 \\
\api{DropAction} & \api{IgnoreAction} & 0 \\
\api{DropAction} & \api{CopyAction} & 1 \\
\api{DropAction} & \api{MoveAction} & 2 \\
\api{DropAction} & \api{LinkAction} & 4 \\
\api{EditTrigger} & \api{NoEditTriggers} & 0 \\
\api{EditTrigger} & \api{CurrentChanged} & 1 \\
\api{EditTrigger} & \api{DoubleClicked} & 2 \\
\api{EditTrigger} & \api{SelectedClicked} & 4 \\
\api{EditTrigger} & \api{EditKeyPressed} & 8 \\
\api{EditTrigger} & \api{AnyKeyPressed} & 16 \\
\api{EditTrigger} & \api{AllEditTriggers} & 31 \\
\bottomrule
\end{longtable}
}

\subsection{Rendering and Image Enums}

{\small
\begin{longtable}{@{}p{0.34\textwidth}p{0.42\textwidth}p{0.12\textwidth}@{}}
\toprule
Enum & Name & Value \\
\midrule
\api{AspectRatioMode} & \api{Ignore} & 0 \\
\api{AspectRatioMode} & \api{Keep} & 1 \\
\api{AspectRatioMode} & \api{KeepByExpanding} & 2 \\
\api{TransformationMode} & \api{Fast} & 0 \\
\api{TransformationMode} & \api{Smooth} & 1 \\
\api{PenStyle} & \api{NoPen} & 0 \\
\api{PenStyle} & \api{SolidLine} & 1 \\
\api{PenStyle} & \api{DashLine} & 2 \\
\api{PenStyle} & \api{DotLine} & 3 \\
\api{PenStyle} & \api{DashDotLine} & 4 \\
\api{PenStyle} & \api{DashDotDotLine} & 5 \\
\api{PenStyle} & \api{CustomDashLine} & 6 \\
\api{PenCapStyle} & \api{FlatCap} & 0 \\
\api{PenCapStyle} & \api{SquareCap} & 16 \\
\api{PenCapStyle} & \api{RoundCap} & 32 \\
\api{PenJoinStyle} & \api{MiterJoin} & 0 \\
\api{PenJoinStyle} & \api{BevelJoin} & 64 \\
\api{PenJoinStyle} & \api{RoundJoin} & 128 \\
\api{PenJoinStyle} & \api{SvgMiterJoin} & 256 \\
\api{PainterCompositionMode} & \api{SourceOver} & 0 \\
\api{PainterCompositionMode} & \api{DestinationOver} & 1 \\
\api{PainterCompositionMode} & \api{Clear} & 2 \\
\api{PainterCompositionMode} & \api{Source} & 3 \\
\api{PainterCompositionMode} & \api{Destination} & 4 \\
\api{PainterCompositionMode} & \api{SourceIn} & 5 \\
\api{PainterCompositionMode} & \api{DestinationIn} & 6 \\
\api{PainterCompositionMode} & \api{SourceOut} & 7 \\
\api{PainterCompositionMode} & \api{DestinationOut} & 8 \\
\api{PainterCompositionMode} & \api{SourceAtop} & 9 \\
\api{PainterCompositionMode} & \api{DestinationAtop} & 10 \\
\api{PainterCompositionMode} & \api{Xor} & 11 \\
\api{PainterCompositionMode} & \api{Plus} & 12 \\
\api{PainterCompositionMode} & \api{Multiply} & 13 \\
\api{PainterCompositionMode} & \api{Screen} & 14 \\
\api{PainterCompositionMode} & \api{Overlay} & 15 \\
\api{PainterCompositionMode} & \api{Darken} & 16 \\
\api{PainterCompositionMode} & \api{Lighten} & 17 \\
\api{PainterCompositionMode} & \api{ColorDodge} & 18 \\
\api{PainterCompositionMode} & \api{ColorBurn} & 19 \\
\api{PainterCompositionMode} & \api{HardLight} & 20 \\
\api{PainterCompositionMode} & \api{SoftLight} & 21 \\
\api{PainterCompositionMode} & \api{Difference} & 22 \\
\api{PainterCompositionMode} & \api{Exclusion} & 23 \\
\bottomrule
\end{longtable}
}

\section{Keeping This Appendix Current}

When a new wrapper enum becomes visible in examples, callbacks, or public setters, add it here if a reader would otherwise need to jump into the source to understand a small code sample. Internal or rarely used enums can stay in the API source until they become part of the guide's narrative.
